---
title: "fiber_LCAT_ER"
author: "Eddie Romo"
date: '2022-08-10'
output:
  html_document:
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

```{r setup, include=FALSE}

if(!any(rownames(installed.packages()) == "pacman")){
   install.packages("pacman")
}

pacman::p_load(here,
               readxl,
               tidyverse,
               lme4,
               lmerTest,
               emmeans,
               car,
               magrittr,
               dplyr,
               readxl,
               ggplot2
)



```

# ID mapping + clinic metadata

```{r}
mapping <- read.csv(here("subject_characteristics","final_samples.csv"))[,1:2] %>%
    `colnames<-`(c("Sample", "Sample.ID")) %>%
    mutate(Sample = as.character(Sample))

clinic <- read_xlsx(here("subject_characteristics","USANA Study Clinical Meta Data Sheet.xlsx"), sheet = 1) %>%
    mutate(
        Sample.ID = paste0(subject, `timepoint coded`),
        treatment = ifelse(treatment=="A", "fiber", "placebo")
    ) 

metadata <- left_join(mapping, clinic)
```

# Plate layout

```{r}
layout1 <- c(as.matrix(read_xlsx(here("lcat","plate_layout","LCAT_layout.xlsx"), range = "B1:L9"))) %>%
    sub(" *", "", .)

layout2 <- c(as.matrix(read_xlsx(here("lcat","plate_layout","LCAT_layout.xlsx"), range = "B12:L20"))) %>%
    sub(" *", "", .)

layout1 <- layout1[!is.na(layout1)]
layout1[layout1=="ercc"] <- "ercc.plate1"
layout1[layout1=="blank"] <- "blank.plate1"

layout2 <- layout2[!is.na(layout2)]
layout2[layout2=="ercc"] <- "ercc.plate2"
layout2[layout2=="blank"] <- "blank.plate2"
```

# Read and tidy LCAT assay results

```{r}
ranges <- c("C145:M153", "C157:M165", "C169:M177")

plate1 <- lapply(ranges, function(x){
    c(as.matrix(read_xlsx(here("2022-07-01_LCAT Fiber.xlsx"), range = x)))
}) %>%
    `names<-`(c("ratio1", "ratio2", "ratio3")) %>%
    do.call(cbind, .) %>%
    as.data.frame() %>%
    filter(complete.cases(.)) %>%
    mutate(Sample = layout1,
           Batch = "1")

plate2 <- lapply(ranges, function(x){
    c(as.matrix(read_xlsx(here("2022-07-02_LCAT Fiber.xlsx"), range = x)))
}) %>%
    `names<-`(c("ratio1", "ratio2", "ratio3")) %>%
    do.call(cbind, .) %>%
    as.data.frame() %>%
    filter(complete.cases(.)) %>%
    mutate(Sample = layout2, 
           Batch = "2")

# Calculate mean LCAT across ratios
plate1 <- plate1 %>% 
        rowwise() %>%
        mutate(mean_ratio = mean(c(ratio1, ratio2, ratio3))) %>%
        ungroup() %>%
        group_by(Sample, Batch) %>%
        summarize(ratio = mean(mean_ratio)) %>%
        ungroup %>%
        mutate(norm_ratio = ratio/ratio[Sample == "ercc.plate1"])

plate2 <- plate2 %>% 
        rowwise() %>%
        mutate(mean_ratio = mean(c(ratio1, ratio2, ratio3))) %>%
        ungroup() %>%
        group_by(Sample, Batch) %>%
        summarize(ratio = mean(mean_ratio)) %>%
        ungroup %>%
        mutate(norm_ratio = ratio/ratio[Sample == "ercc.plate2"])       

plate.merge <- rbind(plate1, plate2) %>%
    left_join(metadata) %>%
    pivot_longer(cols = ratio, names_to = "Read", values_to = "LCAT.ratio") %>%
        pivot_longer(cols = norm_ratio,  values_to = "norm.LCAT.ratio") %>%
    mutate_at(c("Sample", "Batch", "Sample.ID", "subject", "timepoint coded", "gender", "Read"), factor) %>%
    mutate(timepoint = factor(timepoint, levels = c("pre", "post")),
           treatment = factor(treatment, levels = c("placebo", "fiber")))

# Exclude POS, NEG in the mixed linear model
plate.merge.sub <- filter(plate.merge, complete.cases(plate.merge)) %>%
    mutate(EU = paste(subject, timepoint, treatment, sep = "."))
```

# Statistical analysis

EU=treatment:timepoint:subject.\
Linear fixed model: LCAT.ratio\~treatment\*timepoint+(1\|EU)+subject

```{r}
# linear mixed model
lmmod <- lmerTest::lmer(norm.LCAT.ratio~treatment*timepoint+(1|subject) + gender, data = plate.merge.sub)
# Check the assumption: normality
qqPlot(resid(lmmod))
# ANOVA
summary(lmmod)
anova(lmmod, ddf="K")
# post hoc
emm <- emmeans(lmmod, specs = c("treatment", "timepoint"), lmer.df = "K")
con <- contrast(emm, interaction = "pairwise", lmer.df = "K")
con
coef(con)
```

Results showed: (fiber.post-fiber.pre)-(placebo.post-placebo.pre)=0.00867, p-value = 0.5387\
Subjects contributed more variances than treatment and timepoint.

# Visulization

```{r}
# POS, NEG values
ref <- filter(plate.merge, !complete.cases(plate.merge))[,c("Sample", "Batch", "Read", "LCAT.ratio")] %>%
    group_by(Sample, Batch) %>%
    summarise(mean = mean(LCAT.ratio))
```

## All data points

```{r}
# boxplot, all data points
ggplot(plate.merge.sub, aes(timepoint, norm.LCAT.ratio))+
    geom_boxplot()+
    geom_point(aes(color = subject), position = position_jitter())+
    # geom_line(aes(color = subject, group = subject)) +
    facet_grid(~treatment) +
    theme_bw()
```

## Mean for each subject

```{r}
# boxplot, mean for each subject
plate.merge.sub %>%
    ggplot(aes(timepoint, norm.LCAT.ratio))+
    geom_boxplot()+
    geom_point(aes(color = subject))+
    geom_line(aes(color = subject, group = subject)) +
    facet_grid(~treatment) +
    theme_bw()
```

## Mean for each subject + batch 1&2

```{r}
# boxplot, show batch 1&2
plate.merge.sub %>%
    ggplot(aes(timepoint, norm.LCAT.ratio))+
    geom_boxplot()+
    geom_point(aes(color = subject))+
    geom_line(aes(color = subject, group = subject)) +
    facet_grid(Batch~treatment) +
    theme_bw()
```

## Mean for each subject + batch 1&2 + POS + NEG

The positive and negative controls are shown as red horizontal lines. Looks like the negative control in batch and positive control in batch 1 didn't work well?

```{r}
# boxplot, show batch 1&2 and POS, NEG
plate.merge.sub %>%
    group_by(subject, timepoint, treatment, Batch) %>%
    summarise(LCAT.ratio.mean = mean(LCAT.ratio), LCAT.ratio.sd = sd(LCAT.ratio)) %>%
    ggplot(aes(timepoint, LCAT.ratio.mean))+
    geom_boxplot()+
    geom_point(aes(color = subject))+
    geom_line(aes(color = subject, group = subject)) +
    facet_grid(Batch~treatment, scales = "free_y") +
    geom_hline(data=ref, aes(yintercept = mean), color = "red") +
    theme_bw()
```

