---
title: "LBP and LCAT Analysis - Fiber Study"
author: "Brian"
date: "2023-05-17"
output: 
    html_document:
        toc: true
        toc_float: true
        code_folding: hide
---

## Objective

To determine whether LBP concentration and LCAT activity changed from fiber supplementation in healthy participants and to explore if LPB concentration and LCAT activity is correlated with subject characteristics and microbes.

LCAT was normalized to QC plasma pool.

***

## Statistical Analysis

The statistical analyses for this study were conducted using R software (version 4.2.2), a programming language and free software environment for statistical computing and graphics supported by the R Foundation for Statistical Computing.

For the analysis of repeated measures data, a linear mixed model approach was employed. This method is particularly suitable for repeated measures data as it accounts for the within-subject correlation that arises from repeated measurements on the same subject over time. To implement this model, the lmerTest package in R was used, which provides p-values in type III analysis of variance tables with Satterthwaite's or Kenward-Roger's methods for denominator degrees-of-freedom approximation.

Post-hoc tests on the linear mixed models were performed using the emmeans package. This package provided estimated marginal means (EMMs, previously known as least-squares means in the context of traditional regression models) for many linear, generalized linear, and mixed models.

To assess the relationships between variables, a correlation analysis was performed. Because our data involved repeated observations on the same subjects, a standard correlation analysis was not appropriate as it would not account for the non-independence of the observations. Instead, a repeated measures correlation analysis was performed using the rmcorr package in R. This method provides a measure of the common association among repeated measures for two variables, which is particularly useful when the measures are collected on the same metric and over the same timeframe.

All statistical tests were two-tailed, and a p-value of less than 0.05 was considered statistically significant. This threshold was chosen to control the Type I error rate and to balance between the risk of false positives and the risk of false negatives.

***

## Shortened statistical analysis method for manuscript

All analyses were performed using R version 4.2.2. A linear mixed model was implemented with the lmerTest package to account for repeated measures. Post-hoc tests on these models utilized the emmeans package. Correlations between variables, considering the repeated measures nature of the data, were analyzed using the rmcorr package and adjusted for multiple testing using the Benjamini-Hochberg method. Statistical significance was set at a p-value of less than 0.05.

***

## Packages
```{r}
if(!any(rownames(installed.packages()) == "pacman"))
        install.packages("pacman")
if(!any(rownames(installed.packages()) == 'HTSet'))
        devtools::install_github("zhuchcn/HTSet")


pacman::p_load(here,
               tidyverse,
               rstatix,
               DT,
               ggpubr,
               readxl,
               reshape2,
                forcats,
               drc,
               emmeans,
               lmerTest,
               rmcorr,
               HTSet,
               cowplot,
               patchwork,
               RColorBrewer,
               ggsignif,
               viridis,
               ggsci,
               scales,
               openxlsx)
```


## Data Cleaning
```{r message = F}
lcat <- readRDS(here("LCAT","data","LCAT_summary_filter_223B.rds")) 
standard <- read_excel(here("LBP","standard","standard_lbp.xlsx"))
subject_characteristic <- read_csv(here("subject_characteristics","USANA Study Clinical Meta Data Sheet copy.csv")) |>
    mutate(sample = as.character(sample)) |>
    na.omit()
plate_1 <- read_excel(here("LBP","raw","fslbp1-40.xlsx"), 
                    sheet = "Sheet1",
                    range = "B46:N54",
                    
) |>
        column_to_rownames(var = 1) |>
        as.matrix() |>
        melt(value.name = "absorbance",
             na.rm = TRUE)
plate_1$position <- paste0(plate_1$Var1, plate_1$Var2)

        
plate_2 <- read_excel(here("LBP","raw","fslbp41-80.xlsx"), 
                    sheet = "Sheet1",
                    range = "B77:N85",
                    
) |>
        column_to_rownames(var = 1) |>
        as.matrix() |>
        melt(value.name = "absorbance",
             na.rm = TRUE)
plate_2$position <- paste0(plate_2$Var1, plate_2$Var2)

plate_layout_1 <-  read_excel(here("LBP","plate_design","fslbp1-40_plate_design.xlsx"),
                    range = "A1:M9",
                    sheet = "Sheet1"
                    
) |>
        column_to_rownames(var = 1) |>
        as.matrix() |>
        melt() |>
        rename(sample = "value") |>
        filter(!sample %in% "NA")
plate_layout_1$position <- paste0(plate_layout_1$Var1, plate_layout_1$Var2) 


plate_layout_2 <-  read_excel(here("LBP","plate_design","fslbp41-80_plate_design.xlsx"),
                    range = "A1:M9",
                    sheet = "Sheet1"
                    
) |>
        column_to_rownames(var = 1) |>
        as.matrix() |>
        melt() |>
        rename(sample = "value") |>
        filter(!sample %in% "NA")
plate_layout_2$position <- paste0(plate_layout_2$Var1, plate_layout_2$Var2) 

plate_1 <- plate_1 |>
    left_join(plate_layout_1) |>
    mutate(sample = gsub(" ", "",sample))

plate_2 <- plate_2 |>
    left_join(plate_layout_2)
```


## LBP Standard: Four Parameter Logistic Regression
```{r message = F}

plate_1_standard <- plate_1 |>
    left_join(standard)

model1<-drm(absorbance~conc_pg_ml,
             fct=LL.4(names=c("Slope", "Lower", "Upper", "ED50")),
data=plate_1_standard)
plot(model1,
     main = "Plate 1")


plate_2_standard <- plate_2 |>
    left_join(standard)

model2<-drm(absorbance~conc_pg_ml,
             fct=LL.4(names=c("Slope", "Lower", "Upper", "ED50")),
data=plate_2_standard)
plot(model2,
     main = "Plate 2")

```

> For most ELISA, often linear regression does not work. An appropriate curve is usually a four parameter logistic regression.

## Estimate Concentration

```{r message = F}

### Plate 1
dilution_factor <- 800

plate_1_filter <- plate_1 |>
    filter(!grepl(pattern = "S", x = sample))

conc <-as.data.frame(ED(model1,plate_1_filter$absorbance,type="absolute",display=F))
conc_pg_ml <- conc$Estimate * dilution_factor

plate_1_filter <- cbind(plate_1_filter, conc_pg_ml)

### Plate 2

plate_2_filter <- plate_2 |>
    filter(!grepl(pattern = "S", x = sample))

conc <-as.data.frame(ED(model2,plate_2_filter$absorbance,type="absolute",display=F))
conc_pg_ml <- conc$Estimate * dilution_factor

plate_2_filter <- cbind(plate_2_filter, conc_pg_ml)

plate_combined <- plate_1_filter |>
    bind_rows(plate_2_filter) |>
    mutate(conc_ng_ml = conc_pg_ml/1000) |>
    left_join(subject_characteristic) |>
    mutate(treatment = ifelse(treatment=="A", "Prebiotic", "Placebo"),
           treatment = factor(treatment, levels = c("Placebo","Prebiotic")),
           timepoint = factor(timepoint, levels = c("pre", "post")))

plate_summary <- plate_combined |>
    group_by(subject, treatment, timepoint, gender) |>
    summarise(mean_conc_ng_ml = mean(conc_ng_ml),
              sd = sd(conc_ng_ml),
              cv = sd/mean_conc_ng_ml*100) |>
    mutate(subject = as.factor(subject))

datatable(plate_summary, colnames = c("Subject","Treatment","Timepoint","Gender","LBP (ng/mL)","SD","CV (%)")) |>
    formatRound(columns = c("mean_conc_ng_ml","sd","cv"), digits = 2)
```


## Mixed Linear Model

```{r message = F}

mix_model <- lmer(log(mean_conc_ng_ml) ~ treatment*timepoint + (1|subject), data = plate_summary)

# Check the assumption: normality
shapiro.test(resid(mix_model))
# ANOVA
summary(mix_model)
anova(mix_model, ddf="K")
# post hoc
emm <- emmeans(mix_model, specs = c("treatment", "timepoint"), lmer.df = "K")
con <- contrast(emm, interaction = "pairwise", lmer.df = "K")
con
summary_con <- summary(con)
p_value <- round(summary_con$p.value,3)
```

>  The estimated difference between fiber and placebo at the "post" timepoint compared to the "pre" timepoint is -0.0906, but it is not statistically significant (p-value = 0.446), indicating that there is no significant difference between fiber and placebo in their effect on the response variable at different timepoints.

## Table Summary
```{r}
treatment_summary <- plate_combined |>
    group_by(treatment, timepoint) |>
    summarise(mean_conc_ng_ml = round(mean(conc_ng_ml),2),
              sd = round(sd(conc_ng_ml),2))

datatable(treatment_summary, colnames = c("Treatment","Timepoint","LBP (ng/mL)","SD")) |>
    formatRound(columns = c("mean_conc_ng_ml","sd"), digits = 2)

#write_csv(treatment_summary, here("output","LBP_treatment_summary.csv"))

## Pivot Wider
treatment_summary_wide <- treatment_summary |>
    mutate("LBP (ng/mL) +/- SD" = paste0(round(mean_conc_ng_ml,2), " ± ",round(sd,2))) |>
    dplyr::select(-mean_conc_ng_ml,-sd) |>
    pivot_wider(names_from = "timepoint", values_from = "LBP (ng/mL) +/- SD")

```

## LBP Figure
```{r message = F}
# boxplot, all data points
LBP_plot <- ggplot(plate_summary, aes(timepoint, mean_conc_ng_ml))+
    geom_boxplot()+
    geom_point(aes(color = subject))+
    geom_line(aes(color = subject, group = subject)) +
    facet_grid(~treatment) +
    theme_bw() +
        labs(y = "LBP (ng/mL)", x = "Timepoint", title = paste0("P = ",p_value))

LBP_plot

ggsave(plot = LBP_plot, filename  = here("figures","LPB_plot.jpg"),
       dpi = 300,
       width = 8,
       height = 6)
```

> No statistically significant difference in LBP between pre vs post among treatments. P-value is provided on the log scale (p = 0.446)


## LBP Separated by Median Pre LBP
```{r}
# Calculate mean conc_ng_ml for each subject, timepoint, and treatment
plate_responder <- plate_combined %>%
  group_by(subject, treatment, timepoint) %>%
  summarise(mean_conc_ng_ml = mean(conc_ng_ml), .groups = "drop") %>%
  pivot_wider(names_from = timepoint, values_from = mean_conc_ng_ml)

# Calculate change in LBP for each subject and classify as responder or non-responder
plate_responder <- plate_responder %>%
  mutate(change_in_LBP = post - pre,
         response = ifelse(change_in_LBP < median(change_in_LBP, na.rm = TRUE), "responder", "non-responder"))

# Subset to fiber treatment group
plate_responder_fiber <- plate_responder %>%
  filter(treatment == "Prebiotic")

# Calculate median LBP for grouping
median_LBP <- median(plate_responder_fiber$pre, na.rm = TRUE)

# Classify as above or below median LBP
plate_responder_fiber <- plate_responder_fiber %>%
  mutate(group = ifelse(pre < median_LBP, "below median", "above median"))

# Convert to long format for plotting
plate_long_fiber <- plate_responder_fiber %>%
  pivot_longer(cols = c(pre, post), names_to = "timepoint", values_to = "conc_ng_ml") %>%
  mutate(subject = as.factor(subject),
         timepoint = as.factor(timepoint),
         response = as.factor(response),
         group = as.factor(group),
         timepoint = factor(timepoint, levels = c("pre", "post")))





# Create the plot
p <- ggplot(plate_long_fiber, aes(x = timepoint, y = conc_ng_ml, fill = group)) +
  geom_boxplot() +
  geom_point(aes(color = subject)) +
  geom_line(aes(group = subject, color = subject)) +
  facet_grid(.~ group) +
  labs(x = "Timepoint", y = "LBP (ng/mL)", fill = "Group", color = "Subject") +
  theme_minimal() +
  theme(legend.position = "bottom")

p <- ggplot(plate_long_fiber, aes(timepoint, conc_ng_ml))+
    geom_boxplot()+
    geom_point(aes(color = subject))+
    geom_line(aes(color = subject, group = subject)) +
    facet_grid(~group) +
    theme_bw() +
        labs(y = "LBP (ng/mL)", x = "Timepoint")


# Add significance indicators separately for each response type and group
LBP_responder <- p +
  geom_signif(data = subset(plate_long_fiber, group == "above median"), 
              aes(x = timepoint, y = conc_ng_ml), 
              comparisons = list(c("pre", "post")), 
              test = "t.test", 
              test.args = list(paired = TRUE),
              map_signif_level = TRUE) +
  geom_signif(data = subset(plate_long_fiber, group == "below median"), 
              aes(x = timepoint, y = conc_ng_ml), 
              comparisons = list(c("pre", "post")), 
              test = "t.test", 
              test.args = list(paired = TRUE),
              map_signif_level = TRUE)

LBP_responder

# Calculate p-values for each group
p_values_above_median <- with(subset(plate_long_fiber, group == "above median"), t.test(conc_ng_ml ~ timepoint, paired = TRUE)$p.value)
p_values_below_median <- with(subset(plate_long_fiber, group == "below median"), t.test(conc_ng_ml ~ timepoint, paired = TRUE)$p.value)

# Create a table with the p-values
p_value_table <- data.frame(
  group = c("above median", "below median"),
  P_Value = c(p_values_above_median, p_values_below_median)
)
plate_long_fiber_summary <- plate_long_fiber |>
    group_by(group, timepoint) |>
    summarize(mean_conc_ng_ml = mean(conc_ng_ml),
              sd = sd(conc_ng_ml))


# Print the table
datatable(plate_long_fiber_summary)
datatable(p_value_table)

plate_long_fiber_summary_output <- plate_long_fiber_summary |>
    mutate("LBP (ng/mL) +/- SD" = paste0(round(mean_conc_ng_ml,2), " ± ",round(sd,2))) |>
    dplyr::select(-mean_conc_ng_ml,-sd) |>
    pivot_wider(names_from = "timepoint", values_from = "LBP (ng/mL) +/- SD") |>
    left_join(p_value_table)


#write_excel(plate_long_fiber_summary_output, here("output","LBP_median_summary.xlsx"))


ggsave(LBP_responder, filename = here("figures","separate_LBP_median.jpg"),
       dpi = 300,
       width = 8,
       height = 6,
       bg = 'white')


```

> Individuals with above-median pre-LBP levels showed a significant decrease in post-LBP levels after fiber treatment (P < 0.020).


***

### Summary of LBP Responder vs non responder (Fiber Only)
```{r}
# Calculate mean and standard deviation of conc_ng_ml for each combination of response and timepoint
summary_table <- plate_long_fiber %>%
  group_by(response, timepoint) %>%
  summarise(
    mean_conc_ng_ml = mean(conc_ng_ml, na.rm = TRUE),
    sd_conc_ng_ml = sd(conc_ng_ml, na.rm = TRUE)
  )

# Perform paired t-tests and extract p-values
p_values <- plate_long_fiber %>%
  group_by(response) %>%
  summarise(p_value = t.test(conc_ng_ml[timepoint == "pre"], conc_ng_ml[timepoint == "post"], paired = TRUE)$p.value)

# Merge p-values with the summary table
result_table <- left_join(summary_table, p_values, by = "response")
# Round values to three decimal places
result_table$mean_conc_ng_ml <- format(result_table$mean_conc_ng_ml, nsmall = 3)
result_table$sd_conc_ng_ml <- format(result_table$sd_conc_ng_ml, nsmall = 3)
result_table$p_value <- format(result_table$p_value, nsmall = 3)
# Create datatable
datatable(result_table,
          options = list(pageLength = 10, lengthChange = FALSE),
          colnames = c("Response", "Timepoint", "Mean conc_ng_ml", "SD conc_ng_ml", "p-value"))

```

> No difference in groups that had fiber treatment first. 


## LBP: Fiber treated first only
```{r}
# Filter the data for fiber treatment and selected subjects
plate_fiber_first <- plate_combined %>%
  filter(treatment == "Prebiotic" & subject %in% c("201", "203", "205", "209", "213", "216", "222", "223", "224")) |>
    mutate(subject = as.factor(subject))
             


# Create a separate boxplot plot
boxplot_plot <- plate_fiber_first |>
    group_by(subject, treatment, timepoint) |>
    summarize(mean_conc_ng_ml = mean(conc_ng_ml),
              sd = sd(conc_ng_ml)) |>
ggplot(aes(x = timepoint, y = mean_conc_ng_ml, fill = timepoint)) +
  geom_boxplot() +
  geom_point(aes(color = subject)) +
    geom_line(aes(group = subject, color = subject)) +
  geom_signif(comparisons = list(c("pre", "post")), test = "t.test", map_signif_level = TRUE, step_increase = 0.05) +
  labs(x = "Timepoint", y = "LBP Concentration", fill = "Timepoint", color = "Subject") +
  theme_minimal()
boxplot_plot

ggsave(boxplot_plot, filename = here("figures","fiber_first_lbp.jpg"),
       dpi = 300,
       width = 8,
       height = 6)

```

> No significant difference when comparing "Fiber" supplemented first participants

## LCAT activity Table and Figure
```{r message = F}
lcat_rename <- lcat |>
    mutate(treatment = ifelse(treatment=="fiber", "Prebiotic", "Placebo"))
mix_model_lcat <- lmer(norm.LCAT.ratio ~ treatment*timepoint + (1|subject), data = lcat_rename)

# Check the assumption: normality
#shapiro.test(resid(mix_model_lcat))
# post hoc
emm <- emmeans(mix_model_lcat, specs = c("treatment", "timepoint"), lmer.df = "K")
con <- contrast(emm, interaction = "pairwise", lmer.df = "K")
summary_con <- summary(con)
p_value_lcat <- round(summary_con$p.value,3)

lcat_summary <- lcat_rename |>
    group_by(subject, treatment, timepoint) |>
    summarise(mean_norm.LCAT.ratio = mean(norm.LCAT.ratio)) |>
    mutate(treatment = factor(treatment, levels = c("Placebo","Prebiotic")),
           timepoint = factor(timepoint, levels = c("pre", "post")),
           subject = as.factor(subject))

lcat_treatment_summary <- lcat_rename |>
    group_by(subject, treatment, timepoint) |>
    summarise(mean_norm.LCAT.ratio = mean(norm.LCAT.ratio),
              sd = sd(norm.LCAT.ratio)) |>
    mutate(treatment = factor(treatment, levels = c("Placebo","Prebiotic")),
           timepoint = factor(timepoint, levels = c("pre", "post")),
           subject = as.factor(subject))

datatable(lcat_treatment_summary, colnames = c("Subject","Treatment","Timepoint","LCAT Activity Index","SD")) |>
    formatRound(columns = c("mean_norm.LCAT.ratio","sd"), digits = 3)

# boxplot, all data points
LCAT_plot <- ggplot(lcat_summary, aes(timepoint, mean_norm.LCAT.ratio))+
    geom_boxplot()+
    geom_point(aes(color = subject))+
    geom_line(aes(color = subject, group = subject)) +
    facet_grid(~treatment) +
    theme_bw() +
        labs(y = "LCAT Activity Index", x = "Timepoint", title = paste0("P = ",p_value_lcat))

LCAT_plot

ggsave(plot = LCAT_plot, filename = here("figures","LCAT_plot.tiff"),
                                dpi = 300,
       width = 8,
       height = 6)
```

> LCAT activity was not significantly different between pre vs post among treatment groups.

## LCAT output prep to excel
```{r}
lcat_treatment_summary_wide <- lcat_treatment_summary |>
    mutate("LCAT Activity Index +/- SD" = paste0(round(mean_norm.LCAT.ratio,3), " ± ",round(sd,3))) |>
    dplyr::select(-mean_norm.LCAT.ratio,-sd) |>
    pivot_wider(names_from = "timepoint", values_from = "LCAT Activity Index +/- SD")

```



## Write Excel Sheet for LBP Treatment and Median, LCAT Activity Index
```{r}
list_of_datasets <- list("LBP by Treatment" = treatment_summary_wide, 
                         "LBP by Median" = plate_long_fiber_summary_output,
                         "LCAT Activity Index" = lcat_treatment_summary_wide)
write.xlsx(list_of_datasets, file = here("output","LBP Analysis by Treatment and Median, LCAT Activity Index by Treatment.xlsx"))
```

***

## LBP and LCAT Plot
```{r}
# Extract the color legend
common_legend <- get_legend(LCAT_plot + theme(legend.position="bottom"))

# Remove color legends from both plots
LBP_plot <- LBP_plot + theme(legend.position = "none")
LCAT_plot <- LCAT_plot + theme(legend.position = "none")

# Combine the plots vertically
combined_plots <- plot_grid(LBP_plot, LCAT_plot, labels = c("A", "B"), ncol = 2)

# Combine the plots with the common legend
final_plot <- plot_grid(combined_plots, common_legend, ncol = 1, rel_heights = c(1, 0.25))

print(final_plot)
ggsave(plot = final_plot, filename = here("figures","LBP_LCAT_plot.jpg"),
       dpi = 300,
       height = 6,
       width = 8,
       bg = 'white')
```


## LCAT Activity Separated by Median Pre LCAT Activity
```{r}
# Calculate mean norm.LCAT.ratio for each subject, timepoint, and treatment
plate_responder <- lcat_rename %>%
  group_by(subject, treatment, timepoint) %>%
  summarise(mean_norm.LCAT.ratio = mean(norm.LCAT.ratio), .groups = "drop") %>%
  pivot_wider(names_from = timepoint, values_from = mean_norm.LCAT.ratio)

# Calculate change in LCAT for each subject and classify as responder or non-responder
plate_responder <- plate_responder %>%
  mutate(change_in_LCAT = post - pre,
         response = ifelse(change_in_LCAT < median(change_in_LCAT, na.rm = TRUE), "responder", "non-responder"))

# Subset to fiber treatment group
plate_responder_fiber <- plate_responder %>%
  filter(treatment == "Prebiotic")

# Calculate median LCAT Activity for grouping
median_LBP <- median(plate_responder_fiber$pre, na.rm = TRUE)

# Classify as above or below median LCAT Activity 
plate_responder_fiber <- plate_responder_fiber %>%
  mutate(group = ifelse(pre < median_LBP, "below median", "above median"))

# Convert to long format for plotting
plate_long_fiber <- plate_responder_fiber %>%
  pivot_longer(cols = c(pre, post), names_to = "timepoint", values_to = "norm.LCAT.ratio") %>%
  mutate(subject = as.factor(subject),
         timepoint = as.factor(timepoint),
         response = as.factor(response),
         group = as.factor(group),
         timepoint = factor(timepoint, levels = c("pre", "post")))





# Create the plot
p <- ggplot(plate_long_fiber, aes(x = timepoint, y = norm.LCAT.ratio, fill = group)) +
  geom_boxplot() +
  geom_point(aes(color = subject)) +
  geom_line(aes(group = subject, color = subject)) +
  facet_grid(.~ group) +
  labs(x = "Timepoint", y = "LCAT Activity Index", fill = "Group", color = "Subject") +
  theme_minimal() +
  theme(legend.position = "bottom")

p <- ggplot(plate_long_fiber, aes(timepoint, norm.LCAT.ratio))+
    geom_boxplot()+
    geom_point(aes(color = subject))+
    geom_line(aes(color = subject, group = subject)) +
    facet_grid(~group) +
    theme_bw() +
        labs(y = "LCAT Activity Index", x = "Timepoint")


# Add significance indicators separately for each response type and group
LCAT_responder <- p +
  geom_signif(data = subset(plate_long_fiber, group == "above median"), 
              aes(x = timepoint, y = norm.LCAT.ratio), 
              comparisons = list(c("pre", "post")), 
              test = "t.test", 
              test.args = list(paired = TRUE),
              map_signif_level = TRUE) +
  geom_signif(data = subset(plate_long_fiber, group == "below median"), 
              aes(x = timepoint, y = norm.LCAT.ratio), 
              comparisons = list(c("pre", "post")), 
              test = "t.test", 
              test.args = list(paired = TRUE),
              map_signif_level = TRUE)

LCAT_responder

# Calculate p-values for each group
p_values_above_median <- with(subset(plate_long_fiber, group == "above median"), 
                              t.test(norm.LCAT.ratio ~ timepoint, 
                                     paired = TRUE,
                                     na.rm = TRUE)$p.value)

p_values_below_median <- with(subset(plate_long_fiber, group == "below median"),
                              t.test(norm.LCAT.ratio ~ timepoint,
                                     paired = TRUE, 
                                     na.rm = TRUE)$p.value)

# Create a table with the p-values
p_value_table <- data.frame(
  group = c("above median", "below median"),
  P_Value = c(p_values_above_median, p_values_below_median)
)
plate_long_fiber_summary <- plate_long_fiber |>
    group_by(group, timepoint) |>
    summarize(mean_norm.LCAT.ratio = mean(norm.LCAT.ratio, na.rm = TRUE),
              sd = sd(norm.LCAT.ratio, na.rm = TRUE))


# Print the table
datatable(plate_long_fiber_summary)
datatable(p_value_table)

plate_long_fiber_summary_output <- plate_long_fiber_summary |>
    mutate("LCAT Activity Index +/- SD" = paste0(round(mean_norm.LCAT.ratio,3), " ± ",round(sd,2))) |>
    dplyr::select(-mean_norm.LCAT.ratio,-sd) |>
    pivot_wider(names_from = "timepoint", values_from = "LCAT Activity Index +/- SD") |>
    left_join(p_value_table)


#write_excel(plate_long_fiber_summary_output, here("output","LCAT_median_summary.xlsx"))


ggsave(LCAT_responder, filename = here("figures","separate_LCAT_median.jpg"),
       dpi = 300,
       width = 8,
       height = 6,
       bg = 'white')


```


## Combined LCAT and LBP Median
```{r}
# Extract the color legend
common_legend <- get_legend(LCAT_responder + theme(legend.position="bottom"))

# Remove color legends from both plots
LBP_responder <- LBP_responder + theme(legend.position = "none")
LCAT_responder <- LCAT_responder + theme(legend.position = "none")

# Combine the plots vertically
combined_plots <- plot_grid(LBP_responder, LCAT_responder, labels = c("A", "B"), ncol = 2)

# Combine the plots with the common legend
final_plot <- plot_grid(combined_plots, common_legend, ncol = 1, rel_heights = c(1, 0.25))

print(final_plot)
ggsave(plot = final_plot, filename = here("figures","LBP_LCAT_by_MEDIAN_plot.jpg"),
       dpi = 300,
       height = 6,
       width = 8,
       bg = 'white')
```

## Log LBP Correlation with LCAT activity
```{r message = F}
lcat_corr <- lcat_rename |>
    dplyr::select(-Sample, subject, treatment, timepoint, norm.LCAT.ratio) |>
    mutate(subject = as.factor(subject))

lcat_corr <- lcat_corr |>
    left_join(plate_summary)

cor <- rmcorr(subject, measure1 = norm.LCAT.ratio, measure2 = log(mean_conc_ng_ml), dataset = lcat_corr)
cor
```

> Repeated measure correlation shows that LCAT and LBP are not significantly correlated (r = -0.06, P = 0.630).


## Log LBP Repeated Correlation with subject characteristics
```{r warning = F, message = F}
LBP <- log(plate_summary$mean_conc_ng_ml)
# Extract the columns of interest
data_subset <- subject_characteristic[, 9:22]
# Assuming 'subject' is the name of your grouping variable
group_id <- subject_characteristic$subject

# Add the grouping variable and the repeated measures variable to the data subset
data_subset <- cbind(group_id = group_id, LBP = LBP, data_subset)



# List to store the results
results_list_lbp <- list()

# Loop to perform rmcorr for each variable
for(i in 3:(ncol(data_subset))) {
    results <- rmcorr(dataset = data_subset, 
                      participant = "group_id", 
                      measure1 = "LBP", 
                      measure2 = names(data_subset)[i])
    results_list_lbp[[names(data_subset)[i]]] <- results
}


# Convert list to data frame
results_df_lbp <- results_list_lbp %>%
  purrr::map_dfr(function(x) {
    tibble(
      Rmcorr = x$r,
      CI_lower = x$CI[1],
      CI_upper = x$CI[2],
      p_value = x$p,
    padj = p.adjust(p_value, method = "BH"),
    sig = insight::format_p(padj, name = NULL, stars_only = TRUE))
  }, .id = "Variable")

results_df_lbp <- results_df_lbp |>
    mutate(padj = p.adjust(p_value, method = "BH"),
           sig = insight::format_p(padj, name = NULL, stars_only = TRUE))
# Display as data table
datatable(results_df_lbp) |>
    formatRound(columns = c("Rmcorr","CI_lower","CI_upper","p_value","padj"), digits = 3)


```

> The analysis examined the relationship between HDL (high-density lipoprotein) and LBP (lipopolysaccharide-binding protein) concentration using repeated measures correlation. The results showed a significant positive correlation between HDL and LBP concentration (Repeated measure r = 0.271, p = 0.035). However, after adjusting for multiple testing, the association did not remain statistically significant.

***

## LCAT activity Repeated Correlation with subject characteristics
```{r warning = F, message = F}
LCAT <- lcat$norm.LCAT.ratio
# Extract the columns of interest
subject_characteristic_subset <- subject_characteristic |>
    filter(!sample %in% "74")
data_subset <- subject_characteristic_subset[, 9:22] 
    
# Assuming 'subject' is the name of your grouping variable
group_id <- subject_characteristic_subset$subject 
    
    

# Add the grouping variable and the repeated measures variable to the data subset
data_subset <- cbind(group_id = group_id, LCAT = LCAT, data_subset)



# List to store the results
results_list_lcat <- list()

# Loop to perform rmcorr for each variable
for(i in 3:(ncol(data_subset))) {
    results <- rmcorr(dataset = data_subset, 
                      participant = "group_id", 
                      measure1 = "LCAT", 
                      measure2 = names(data_subset)[i])
    results_list_lcat[[names(data_subset)[i]]] <- results
}


# Convert list to data frame

results_df_lcat <- results_list_lcat|>
  purrr::map_dfr(function(x) {
    tibble(
      Rmcorr = x$r,
      CI_lower = x$CI[1],
      CI_upper = x$CI[2],
      p_value = x$p
    )
  }, .id = "Variable")

results_df_lcat <- results_df_lcat |>
    mutate(padj = p.adjust(p_value, method = "BH"),
           sig = insight::format_p(padj, name = NULL, stars_only = TRUE))
# Display as data table
datatable(results_df_lcat) |>
    formatRound(columns = c("Rmcorr","CI_lower","CI_upper","p_value","padj"), digits = 3)

```

> LCAT activity was positively correlated with cholesterol (repeated r = 0.382, p-value = 0.018).

> LCAT activity was positively correlated with HDL (repeated r = 0.391, p-value = 0.018).

> LCAT activity was positively correlated with LDL (repeated r = 0.343, p-value = 0.034).

These results indicate that there is a significant positive correlation between LCAT activity and cholesterol, LCAT activity and HDL, and LCAT activity and LDL. The p-values associated with these correlations indicate the statistical significance of the observed relationships. An adjusted p-value less than the conventional threshold (e.g., 0.05) suggests that the correlation is unlikely to occur by chance.


## Combine LCAT and LBP Correlation with Clinical for Manuscript
```{r}

results_df_lbp <- results_df_lbp |>
    mutate(Rmcorr = round(Rmcorr,2),
           CI_lower = round(CI_lower, 2),
           CI_upper = round(CI_lower, 2),
        `95% CI` = paste0(CI_lower,", ", CI_upper),
.after = Rmcorr) |>
    dplyr::select(-CI_lower,-CI_upper)

results_df_lcat <- results_df_lcat |>
    mutate(Rmcorr = round(Rmcorr,2),
           CI_lower = round(CI_lower, 2),
           CI_upper = round(CI_lower, 2),
        `95% CI` = paste0(CI_lower,", ", CI_upper),
.after = Rmcorr) |>
    dplyr::select(-CI_lower,-CI_upper)

#list_of_datasets_clinical <- list("LBP Rmcorr Clinical" = results_df_lbp, 
                         #"LCAT Rmcorr Clinical" = results_df_lcat)
#write.xlsx(list_of_datasets_clinical, file = here("output","LBP and LCAT Rmcorr with clinical data.xlsx"))

```

***

## LBP Correlation with Microbiome
```{r}
taxa <- readRDS(here("microbiome","data","taxa.rds"))
taxa.id <- paste0(taxa$pdata$subject, tolower(taxa$pdata$trt), taxa$pdata$timepoint)
## Change "fiber" name to Prebiotic and change "placebo" to "Placebo"
taxa.id <- gsub("fiber", "Prebiotic", taxa.id)
taxa.id <- gsub("placebo", "Placebo", taxa.id)

LBP.id <- paste0(plate_summary$subject, plate_summary$treatment, plate_summary$timepoint)
overlapped.id <- taxa.id %in% LBP.id

genus <- summarize_feature(taxa, "genus")[,overlapped.id]

pdata <- genus$pdata # metadata of subjects
genus <- genus$edata %>%
        `colnames<-`(taxa.id[overlapped.id]) # OTU count table at the genus level

var0 <- apply(genus, 1, function(x)var(x)==0) # remove genus with variance == 0

sum(var0)
genus <- genus[!var0, ]

genus.cpm <- apply(genus, 2, function(x){x/sum(x)*1000000}) # total count = 1M
```

```{r}
rmcorrelation <- function(microbe, y, data){
    df <- data.frame(
        subject = factor(pdata$subject),
        microbe = log(data[microbe,]+1), # log-transformation
        LBP = y
    )
    cor <- rmcorr(subject, measure1 = microbe, measure2 = LBP, dataset = df)
}

# repeated measure correlation on all genus
rmcor.res <- lapply(rownames(genus.cpm), function(x)
    rmcorrelation(x, LBP, data = genus.cpm)) %>%
    `names<-`(rownames(genus.cpm))
# extract p-values
pvals <- sapply(rmcor.res, function(x)x$p) %>%
    `names<-`(rownames(genus.cpm))
# padj
padjs <- p.adjust(pvals, method = "BH")

```

> Microbes counts per million (cpm) were log transformed.

### Significantly correlation between log(LBP) ~ log(microbes)
```{r}
rmcor.plot <- function(x, xlab, ylab){
    df <- x$model$model
    p <- ggplot(df, aes(x = Measure1, y = Measure2)) +
        geom_point(aes(colour = factor(Participant), group = factor(Participant))) +
        geom_line(aes(y = x$model$fitted.values, group = factor(Participant), color = factor(Participant)), linetype = 1) +
        theme_bw() +
        labs(x = paste(xlab, "log(counts)"), y = ylab)
    xmax <- max(df$Measure1)
    ymax <- max(df$Measure2)
    p <- p +
        annotate(
            "text", x = xmax*0.93, y = c(ymax, ymax*1.2), 
            label = paste0(c("r = ", "pval = "), signif(c(x$r, x$p), 3))
        )
    return(p)
}
plots <- lapply(rownames(genus.cpm)[which(pvals<0.05)], function(x){
    rmcor.plot(rmcor.res[[x]], xlab = x, ylab = "Log(LBP)")
}) %>%
    `names<-`(rownames(genus.cpm)[which(pvals<0.05)])
plots
```

> 5 Microbes were significantly correlated with LBP (unadjusted p-value). However, none were significant after multiple testing correction. 


## Save the five plots for manuscript
```{r}
# Extract each plot into a separate object
plot1 <- plots[[1]] + labs(color = "subject") + theme(text = element_text(size=20)) + theme(legend.position = "none")
plot2 <- plots[[2]] + labs(color = "subject") + theme(text = element_text(size=20)) + theme(legend.position = "none")
plot3 <- plots[[3]] + labs(color = "subject") + theme(text = element_text(size=20)) + theme(legend.position = "none")
plot4 <- plots[[4]] + labs(color = "subject") + theme(text = element_text(size=20)) + theme(legend.position = "none")
plot5 <- plots[[5]] + labs(color = "subject") + theme(text = element_text(size=20)) + theme(legend.position = "none")


## Extract Common legend
plot1_legend <- plots[[1]] + labs(color = "subject") + theme(text = element_text(size=20)) + guides(color = guide_legend(ncol = 3))

leg <- get_legend(plot1_legend)
leg <- as_ggplot(leg)

figure_LBP_microbiome <- ggarrange(plot1, plot2, plot3, plot4, plot5, leg,
align='h', labels=c('A', 'B','C','D',"E", "F"))


figure_LBP_microbiome <- plot1 + theme(legend.position = "none") +
    plot2 + theme(axis.title.y = element_blank(), legend.position = "none") +
    plot3 + theme(axis.title.y = element_blank(), legend.position = "none") +
    plot4 + theme(legend.position = "none") +
    plot5 + theme(axis.title.y = element_blank(), legend.position = "none") +
    leg + theme(text = element_text(size=18)) +
        plot_layout(ncol = 3, nrow = 2) +
    plot_annotation(tag_levels = 'A')

ggsave(here("figures","LBP_microbiome_corr.jpg"), 
       figure_LBP_microbiome, 
       scale = 2, 
       dpi = 300,
       width = 10,
       height = 6,
       bg = "white")
```

***

### Table of Repeated Measure Correlation (LBP and Microbes)

```{r}
rmcor.cpm <- data.frame(
    microbes = rownames(genus.cpm),
    rmcor = sapply(rmcor.res, function(x)x$r),
    pval = pvals,
    padj = padjs,
    sig = insight::format_p(padjs, name = NULL, stars_only = TRUE)
)
datatable(rmcor.cpm, colnames = c("Genus","Repeated Measure Correlation (rmcorr)","p-value","Adjusted p-value","Sig level")) %>%
        formatRound(columns = "rmcor", digits = 2) %>%
        formatRound(columns = c("pval","padj"), digits = 3)

rmcor.cpm.lbp <- rmcor.cpm
```

## Significance and direction LBP and Microbe
```{r}
nrow(rmcor.cpm)
# Positive Rmcorr
pos_count <- sum(rmcor.cpm$pval[rmcor.cpm$rmcor > 0] < 0.05)
pos_count
# Negative Rmcorr  
neg_count <- sum(rmcor.cpm$pval[rmcor.cpm$rmcor < 0] < 0.05)
neg_count
```

> Of the 185 microbes examined, 3 showed statistically significant positive correlations, while 2 showed statistically significant negative correlations with LBP.

***



## LCAT activity Correlation with Microbiome
```{r}
taxa <- readRDS(here("microbiome","data","taxa.rds"))
taxa.id <- paste0(taxa$pdata$subject, tolower(taxa$pdata$trt), taxa$pdata$timepoint)
LCAT.id <- paste0(lcat$subject, lcat$treatment, lcat$timepoint)
overlapped.id <- taxa.id %in% LCAT.id

genus <- summarize_feature(taxa, "genus")[,overlapped.id]

pdata <- genus$pdata # metadata of subjects
genus <- genus$edata %>%
        `colnames<-`(taxa.id[overlapped.id]) # OTU count table at the genus level

var0 <- apply(genus, 1, function(x)var(x)==0) # remove genus with variance == 0

sum(var0)
genus <- genus[!var0, ]

genus.cpm <- apply(genus, 2, function(x){x/sum(x)*1000000}) # total count = 1M
```

```{r}
rmcorrelation <- function(microbe, y, data){
    df <- data.frame(
        subject = factor(pdata$subject),
        microbe = log(data[microbe,]+1), # log-transformation
        LBP = y
    )
    cor <- rmcorr(subject, measure1 = microbe, measure2 = LCAT, dataset = df)
}

# repeated measure correlation on all genus
rmcor.res <- lapply(rownames(genus.cpm), function(x)
    rmcorrelation(x, LCAT, data = genus.cpm)) %>%
    `names<-`(rownames(genus.cpm))
# extract p-values
pvals <- sapply(rmcor.res, function(x)x$p) %>%
    `names<-`(rownames(genus.cpm))
# padj
padjs <- p.adjust(pvals, method = "BH")

```

> Microbes counts per million (cpm) were log transformed.

### Significantly correlation between LCAT activity Index ~ log(microbes)
```{r}
rmcor.plot <- function(x, xlab, ylab){
    df <- x$model$model
    p <- ggplot(df, aes(x = Measure1, y = Measure2)) +
        geom_point(aes(colour = factor(Participant), group = factor(Participant))) +
        geom_line(aes(y = x$model$fitted.values, group = factor(Participant), color = factor(Participant)), linetype = 1) +
        theme_bw() +
        labs(x = paste(xlab, "log(counts)"), y = ylab)
    xmax <- max(df$Measure1)
    ymax <- max(df$Measure2)
    p <- p +
        annotate(
            "text", x = xmax*0.8, y = c(ymax, ymax*0.99), 
            label = paste0(c("r = ", "pval = "), signif(c(x$r, x$p), 3))
        )
    return(p)
}
plots <- lapply(rownames(genus.cpm)[which(pvals<0.05)], function(x){
    rmcor.plot(rmcor.res[[x]], xlab = x, ylab = "LCAT Activity Index")
}) %>%
    `names<-`(rownames(genus.cpm)[which(pvals<0.05)])
plots
```

> In the analysis, correlations were examined between LCAT (Lecithin-Cholesterol Acyltransferase) activity and 14 different microbes. The initial results showed that all 14 microbes had significant correlations with LCAT, as indicated by unadjusted p-values. However, after applying correction none were statistically significant.

***

## Save the 14 plots for manuscript
```{r}
# Extract each plot into an object
plot1 <- plots[[1]] + labs(color = "subject") + 
   theme(text = element_text(size=18), legend.position = "none")

plot2 <- plots[[2]] + labs(color = "subject") +
   theme(text = element_text(size=18), legend.position = "none")  

plot3 <- plots[[3]] + labs(color = "subject") +
   theme(text = element_text(size=18), legend.position = "none")

plot4 <- plots[[4]] + labs(color = "subject") +
   theme(text = element_text(size=18), legend.position = "none")

plot5 <- plots[[5]] + labs(color = "subject") +
   theme(text = element_text(size=18), legend.position = "none")

plot6 <- plots[[6]] + labs(color = "subject") +
   theme(text = element_text(size=18), legend.position = "none")
plot7 <- plots[[7]] + labs(color = "subject") +
   theme(text = element_text(size=18), legend.position = "none")
plot8 <- plots[[8]] + labs(color = "subject") +
   theme(text = element_text(size=18), legend.position = "none")
plot9 <- plots[[9]] + labs(color = "subject") +
   theme(text = element_text(size=18), legend.position = "none")
plot10 <- plots[[10]] + labs(color = "subject") +
   theme(text = element_text(size=18), legend.position = "none")
plot11 <- plots[[11]] + labs(color = "subject") +
   theme(text = element_text(size=18), legend.position = "none")
plot12 <- plots[[12]] + labs(color = "subject") +
   theme(text = element_text(size=18), legend.position = "none")
plot13 <- plots[[13]] + labs(color = "subject") +
   theme(text = element_text(size=18), legend.position = "none")
plot14 <- plots[[14]] + labs(color = "subject") +
   theme(text = element_text(size=18), legend.position = "none")

## Extract Common legend
plot1_legend <- plots[[1]] + labs(color = "subject") + theme(text = element_text(size=20)) + guides(color = guide_legend(ncol = 3))

leg <- get_legend(plot1_legend)
leg <- as_ggplot(leg)

# Arrange plots in 4x4 grid with legend
figure_LCAT_cor <- ggarrange(plot1, plot2, plot3, plot4,  
                    plot5, plot6, plot7, 
                    plot8, plot9, plot10,
                    plot11, plot12, plot13,
                    plot14,leg,
                    ncol = 5,
                    nrow = 3,
                    common.legend = FALSE)


figure_LCAT_cor <- plot1 + theme(legend.position = "none") +
    plot2 + theme(axis.title.y = element_blank(), legend.position = "none") +
    plot3 + theme(axis.title.y = element_blank(), legend.position = "none") +
    plot4 + theme(axis.title.y = element_blank(), legend.position = "none") +
    plot5 + theme(axis.title.y = element_blank(), legend.position = "none") +
    plot6 + theme(legend.position = "none") +
    plot7 + theme(axis.title.y = element_blank(), legend.position = "none") +
    plot8 + theme(axis.title.y = element_blank(), legend.position = "none") +
    plot9 + theme(axis.title.y = element_blank(), legend.position = "none") +
    plot10 + theme(axis.title.y = element_blank(), legend.position = "none") +
    plot11 + theme(legend.position = "none") +
    plot12 + theme(axis.title.y = element_blank(), legend.position = "none") +
    plot13 + theme(axis.title.y = element_blank(), legend.position = "none") + 
    plot14 +theme(axis.title.y = element_blank(), legend.position = "none") +
    plot_layout(ncol = 5, nrow = 3) +
    leg + theme(text = element_text(size=18)) +
    plot_annotation(tag_levels = 'A')
                    
# Save figure                    
ggsave(figure_LCAT_cor, filename = here("figures","LCAT_microbiome_corr.tiff"),
  scale = 1.5, 
  dpi = 300,
  width = 15, height = 10,
  bg = "white")


```


### Table of Repeated Measure Correlation (Normalized LCAT activity and Microbes)

```{r}
rmcor.cpm <- data.frame(
    microbes = rownames(genus.cpm),
    rmcor = sapply(rmcor.res, function(x)x$r),
    pval = pvals,
    padj = padjs,
    sig = insight::format_p(padjs, name = NULL, stars_only = TRUE)
)
datatable(rmcor.cpm, colnames = c("Genus","Correlation (r)","p-value","Adjusted p-value","Sig level")) %>%
        formatRound(columns = "rmcor", digits = 2) %>%
        formatRound(columns = c("pval","padj"), digits = 3)

rmcor.cpm.LCAT <- rmcor.cpm
```

## Significance and direction LCAT and Microbe
```{r}
nrow(rmcor.cpm)
# Positive Rmcorr
pos_count <- sum(rmcor.cpm$pval[rmcor.cpm$rmcor > 0] < 0.05)
pos_count
# Negative Rmcorr  
neg_count <- sum(rmcor.cpm$pval[rmcor.cpm$rmcor < 0] < 0.05)
neg_count
```

> Of the 185 microbes examined, 11 showed statistically significant positive correlations, while 3 showed statistically significant negative correlations with LCAT activity.

## Write Excel for LBP and LCAT correlaton with Microbes
```{r}
list_of_datasets <- list("LBP corr Microbes" = rmcor.cpm.LBP, 
                         "LCAT corr Microbes" = rmcor.cpm.LCAT)

write.xlsx(list_of_datasets, file = here("output","LBP and LCAT Corr Microbes.xlsx"))
```

***


## Metabolome correlation with LBP

### Data Cleanup
```{r message = F, warning = F}
metabolites <- read_excel(here("metabolomic","raw","UCDA-03-20MD DATA TABLES.xlsx"),
           sheet = "Batch-norm Imputed Data")

sample_meta <- read_excel(here("metabolomic","raw","UCDA-03-20MD DATA TABLES.xlsx"),
           sheet = "Sample Meta Data")

sample_meta_subset <- sample_meta |>
    dplyr::select(PARENT_SAMPLE_NAME, CLIENT_SAMPLE_ID)
chemical_annotation <- read_excel(here("metabolomic","raw","UCDA-03-20MD DATA TABLES.xlsx"),
           sheet = "Chemical Annotation")

chemical_annotation_subset <- chemical_annotation |>
    dplyr::select(CHEM_ID, PLOT_NAME, SUPER_PATHWAY,SUB_PATHWAY) |>
    mutate(CHEM_ID = as.character(CHEM_ID))

plate_mean_lbp <- plate_combined |>
    group_by(subject, treatment, timepoint, `timepoint coded`) |>
    summarise(mean_conc_ng_ml = mean(conc_ng_ml)) |>
    mutate(subject = as.character(subject))

metabolites_full <- metabolites |>
    left_join(sample_meta_subset) |>
    relocate(CLIENT_SAMPLE_ID) |>
  separate(CLIENT_SAMPLE_ID, into = c("subject", "timepoint coded"), sep = "(?<=\\d)(?=\\D)", remove = TRUE) |>
    left_join(plate_mean_lbp) |>
    relocate(treatment, timepoint, mean_conc_ng_ml) 
```


### LBP and Metabolite repeated measure corr
```{r message = F, warning = F}
LBP <- log(metabolites_full$mean_conc_ng_ml)
# Extract the columns of interest
data_subset <- metabolites_full[, 7:896]
# Assuming 'subject' is the name of your grouping variable
group_id <- as.factor(metabolites_full$subject)

# Add the grouping variable and the repeated measures variable to the data subset
data_subset <- cbind(group_id = group_id, LBP = LBP, data_subset)



# List to store the results
results_list <- list()

# Loop to perform rmcorr for each variable
for(i in 3:(ncol(data_subset))) {
    results <- rmcorr(dataset = data_subset, 
                      participant = "group_id", 
                      measure1 = "LBP", 
                      measure2 = names(data_subset)[i])
    results_list[[names(data_subset)[i]]] <- results
}


# Convert list to data frame
results_df <- results_list %>%
  purrr::map_dfr(function(x) {
    tibble(
      Rmcorr = x$r,
      CI_lower = x$CI[1],
      CI_upper = x$CI[2],
      p_value = x$p,
    padj = p.adjust(p_value, method = "BH"),
    sig = insight::format_p(padj, name = NULL, stars_only = TRUE))
  }, .id = "CHEM_ID")

results_df <- results_df |>
    mutate(padj = p.adjust(p_value, method = "BH"),
           sig = insight::format_p(padj, name = NULL, stars_only = TRUE)) |>
    left_join(chemical_annotation_subset) |>
    relocate(PLOT_NAME, SUPER_PATHWAY, SUB_PATHWAY) |>
    dplyr::select(!CHEM_ID) |>
    mutate(SUPER_PATHWAY = as.factor(SUPER_PATHWAY))
# Display as data table
datatable(results_df, colnames = c("Compound","Super Pathway","Subpathway","Rmcorr","CI_lower","CI_upper","p_value","padj","sig")) |>
    formatRound(columns = c("Rmcorr","CI_lower","CI_upper","p_value","padj"), digits = 3)
#sum(results_df$p_value < 0.05)

#write.csv(results_df,here("output","LBP_metabolite_corr.csv"))

```

> There are 50 metabolites out of 890 metabolites were significantly correlated with LBP, but none were signficant after multiple testing. 

## Significance and direction LBP and Plasma Metabolite
```{r}
nrow(results_df)
# Positive Rmcorr
pos_count <- sum(results_df$p_value[results_df$Rmcorr > 0] < 0.05)
pos_count
# Negative Rmcorr  
neg_count <- sum(results_df$p_value[results_df$Rmcorr < 0] < 0.05)
neg_count
```

> Of the 890 plasma metabolites examined, 35 showed statistically significant positive correlations, while 15 showed statistically significant negative correlations with LBP. 

## LBP corr with metabolite barplot
```{r}

# Select the top 30 absolute Rmcorr 
top_results_df <- results_df %>%
    mutate(abs_Rmcorr = abs(Rmcorr)) %>%
    top_n(30, abs_Rmcorr)

# Get a palette of 9 colors
#color_palette <- brewer.pal(9, "Set1")
n_colors <- 9
color_palette <- pal_npg("nrc")(n_colors)


# Create a named vector mapping SUPER_PATHWAY levels to colors ("Energy" and "Peptide" is removed for combined bar plot, copy below  palette when outputting single figure)

#color_mapping <- setNames(color_palette, c("Amino Acid",
#                                           "Carbohydrate", 
 #                                          "Cofactors and Vitamins", #
 #                                          "Energy", 
  #                                         "Lipid", 
   #                                        "Nucleotide", 
    #                                       "Partially Characterized Molecules", 
#                                           "Peptide", 
 #                                          "Xenobiotics"))

color_mapping <- setNames(color_palette, c("Amino Acid",
                                           "Carbohydrate", 
                                           "Cofactors and Vitamins", 
                                            "Lipid", 
                                           "Nucleotide", 
                                           "Partially Characterized Molecules", 
                                           "Xenobiotics"))


# Create the bar plot
top30_lbp_metabolite_plot <- ggplot(top_results_df, aes(x = reorder(PLOT_NAME, abs_Rmcorr), y = Rmcorr, fill = SUPER_PATHWAY)) +
    geom_bar(stat = "identity") +
    coord_flip() +
    labs(x = "Top 30 Compounds", y = "Repeated Measure (rmcorr)", fill = "Super Pathway", title = "LBP") +
    scale_fill_manual(values = color_mapping,
                     drop = FALSE) + # drop = FALSE keeps color level even if not present. Remove this when saving single figure.
    theme_minimal() + 
    theme(legend.position = "bottom",
          plot.title = element_text(hjust = 0.5, color = "black"),
          text = element_text(size = 20, color = "black"),
          axis.text = element_text(color = "black"),
          axis.title = element_text(color = "black"),
          legend.text = element_text(color = "black"),
          legend.title = element_text(color = "black")) +
    guides(fill = guide_legend(nrow = 2))
    

print(top30_lbp_metabolite_plot)
#ggsave(top30_lbp_metabolite_plot, filename = here("figures","LBP_metabolite_corr.jpg"),
 #      dpi= 300,
  #     height = 8,
   #    width = 10,
    #   scale = 1.8)
```

> The top 30 absolute correlation (rrmcorr) is displayed in this bar plot.

***


## Metabolome correlation with LCAT Activity
```{r message = F, warning = F}
metabolites_full_lcat <- lcat_summary |>
    mutate(subject = as.character(subject)) |>
    left_join(metabolites_full)
```

### LCAT Activity and Metabolite repeated measure corr
```{r message = F, warning = F}
LCAT <- metabolites_full_lcat$mean_norm.LCAT.ratio
# Extract the columns of interest
data_subset <- metabolites_full_lcat[, 8:897]
# Assuming 'subject' is the name of your grouping variable
group_id <- as.factor(metabolites_full_lcat$subject)

# Add the grouping variable and the repeated measures variable to the data subset
data_subset <- cbind(group_id = group_id, LCAT = LCAT, data_subset)



# List to store the results
results_list <- list()

# Loop to perform rmcorr for each variable
for(i in 3:(ncol(data_subset))) {
    results <- rmcorr(dataset = data_subset, 
                      participant = "group_id", 
                      measure1 = "LCAT", 
                      measure2 = names(data_subset)[i])
    results_list[[names(data_subset)[i]]] <- results
}


# Convert list to data frame
results_df <- results_list %>%
  purrr::map_dfr(function(x) {
    tibble(
      Rmcorr = x$r,
      CI_lower = x$CI[1],
      CI_upper = x$CI[2],
      p_value = x$p,
    padj = p.adjust(p_value, method = "BH"),
    sig = insight::format_p(padj, name = NULL, stars_only = TRUE))
  }, .id = "CHEM_ID")

results_df <- results_df |>
    mutate(padj = p.adjust(p_value, method = "BH"),
           sig = insight::format_p(padj, name = NULL, stars_only = TRUE)) |>
    left_join(chemical_annotation_subset) |>
    relocate(PLOT_NAME, SUPER_PATHWAY, SUB_PATHWAY) |>
    dplyr::select(!CHEM_ID) |>
    mutate(SUPER_PATHWAY = as.factor(SUPER_PATHWAY))
# Display as data table
datatable(results_df, colnames = c("Compound","Super Pathway","Subpathway","Rmcorr","CI_lower","CI_upper","p_value","padj","sig")) |>
    formatRound(columns = c("Rmcorr","CI_lower","CI_upper","p_value","padj"), digits = 3)

#sum(results_df$p_value < 0.05)

#write.csv(results_df,here("output","LCAT_metabolite_corr.csv"))

```

> 47/890 metabolites were significantly correlated with LCAT activity, however none reach statistical significance after multiple testing.

## Significance and direction LBP and Plasma Metabolite
```{r}
nrow(results_df)
# Positive Rmcorr
pos_count <- sum(results_df$p_value[results_df$Rmcorr > 0] < 0.05)
pos_count
# Negative Rmcorr  
neg_count <- sum(results_df$p_value[results_df$Rmcorr < 0] < 0.05)
neg_count
```

> Of the 890 plasma metabolites examined, 15 showed statistically significant positive correlations, while 18 showed statistically significant negative correlations with LCAT activity. 

## LCAT corr with metabolite barplot
```{r}
# Select the top 30 absolute Rmcorr 
top_results_df <- results_df %>%
    mutate(abs_Rmcorr = abs(Rmcorr)) %>%
    top_n(30, abs_Rmcorr)

# Get a palette of 9 colors

n_colors <- 9
color_palette <- pal_npg("nrc")(n_colors)

#color_palette <- brewer.pal(9, "Set1")

# Create a named vector mapping SUPER_PATHWAY levels to colors
#color_mapping <- setNames(color_palette, c("Amino Acid",
#                                           "Carbohydrate", 
 #                                          "Cofactors and Vitamins", #
 #                                          "Energy", 
  #                                         "Lipid", 
   #                                        "Nucleotide", 
    #                                       "Partially Characterized Molecules", 
#                                           "Peptide", 
 #                                          "Xenobiotics"))


color_mapping <- setNames(color_palette, c("Amino Acid",
                                           "Carbohydrate", 
                                           "Cofactors and Vitamins", 
                                            "Lipid", 
                                           "Nucleotide", 
                                           "Partially Characterized Molecules", 
                                           "Xenobiotics"))

# Create the bar plot
top30_lcat_metabolite_plot <- ggplot(top_results_df, aes(x = reorder(PLOT_NAME, abs_Rmcorr), y = Rmcorr, fill = SUPER_PATHWAY)) +
    geom_bar(stat = "identity") +
    coord_flip() +
    labs(x = "Top 30 Compounds", y = "Repeated Measure (rmcorr)", fill = "Super Pathway", title = "LCAT Activity Index") +
    scale_fill_manual(values = color_mapping) +
    theme_minimal() + 
    theme(legend.position = "bottom",
          plot.title = element_text(hjust = 0.5, color = "black"),
          text = element_text(size = 20, color = "black"),
          axis.text = element_text(color = "black"),
          axis.title = element_text(color = "black"),
          legend.text = element_text(color = "black"),
          legend.title = element_text(color = "black")) +
    guides(fill = guide_legend(nrow = 2))

print(top30_lcat_metabolite_plot)
ggsave(top30_lcat_metabolite_plot, filename = here("figures","LCAT_metabolite_corr.tiff"),
       dpi= 300,
       height = 8,
       width = 10,
       scale = 1.8,
       bg = "white")
```

> The top 30 absolute correlation (rrmcorr) is displayed in this bar plot.


## Combined Metabolite Correlation with LBP and LCAT
```{r, fig.width=20, fig.height=8}


# Combine the plots vertically
#combined_plots <- plot_grid(top30_lbp_metabolite_plot, top30_lcat_metabolite_plot, labels = c("A", "B"), ncol = 2)

combined_plots <- ggarrange(top30_lbp_metabolite_plot, top30_lcat_metabolite_plot, ncol=2, nrow=1, common.legend = TRUE, legend="bottom", labels = "AUTO", font.label = list(size = 30))




ggsave(plot = combined_plots, filename = here("figures","LBP_LCAT_metabolite_plot.tiff"),
       dpi = 300,
       height = 6,
       width = 10,
       scale = 2.5,
       bg = "white")
```

***

## Summary 

In the analysis, several findings were observed:

1) There was no statistically significant difference in LBP (lipopolysaccharide binding protein) levels between the Placebo and control groups at the Pre and Post timepoints.

2) Initially, LBP showed a positive correlation with HDL (high-density lipoprotein) levels, but this correlation did not remain significant after correcting for multiple testing.

3) There was no statistically significant correlation found between LBP and LCAT (lecithin-cholesterol acyltransferase).

4) LBP was found to be significantly correlated with 5 specific microbes, but these correlations did not hold significance after adjusting for multiple testing.

5) LCAT showed significant positive correlations with HDL, LDL (low-density lipoprotein), and cholesterol levels even after adjustment.

6) LCAT was significantly correlated with 14 specific microbes, but these correlations did not remain significant after adjustment.

7) LBP was significantly correlated with 50 metabolites in the unadjusted pvalue, but not adjusted pvalue

8) LCAT Activity was significantly correlated with 47 metabolites in the unadjusted pvalue, but not adjusted pvalue.

9) Individuals with above-median pre-LBP levels showed a significant decrease in post-LBP levels after fiber treatment (P < 0.020).