---
title: "LBP"
format: html
editor: visual
---

## Fiber Study LBP concentration

```{r include=FALSE}
library(tidyverse)
library(readxl)
library(lmerTest)
library(emmeans)
library(multcomp)
library(car)
```

```{r echo=FALSE}
metadata <- read.csv("USANA Study Clinical Meta Data Sheet copy.csv")
metadata$sample <- paste0("SPL", metadata$sample)
metadata$subject <- factor(metadata$subject)
LBP1.40.1 <- read_xlsx("fslbp1-40.xlsx", range = "C46:N54")
LBP1.40.2 <- read_xlsx("fslbp1-40.xlsx", range = "C59:N67")
LBP1.40.3 <- read_xlsx("fslbp1-40.xlsx", range = "C72:N80")
LBP41.80.1 <- read_xlsx("fslbp41-80.xlsx", range = "C51:N59")
LBP41.80.2 <- read_xlsx("fslbp41-80.xlsx", range = "C64:N72")
LBP41.80.3 <- read_xlsx("fslbp41-80.xlsx", range = "C77:N85")
STD = rep(50*0.5^seq(0, 6), each = 2) # 0=blank

# ---------------------Sample 1-40 -------------------------

names <- rep(paste0("SPL", 1:40), each = 2)

LBP1.40 <- list(rep1 = LBP1.40.1, rep2 = LBP1.40.2, rep3 = LBP1.40.3)

# STD tables:

LBP1.40.std <- lapply(names(LBP1.40), function(x){
    read.list <- LBP1.40[[x]] %>% 
        as.matrix() %>% 
        as.numeric()
    blank <- read.list[(length(STD)+1):(length(STD)+2)]
    blank <- mean(blank)
    
    LBP.std <- data.frame(
        std = STD,
        read = read.list[1:length(STD)]-blank,
        replicate = x
    )
    
}) %>%
    Reduce(rbind, .)

# Sample tables:

LBP1.40.sample <- lapply(names(LBP1.40), function(x){
    read.list <- LBP1.40[[x]] %>% 
        as.matrix() %>% 
        as.numeric()
    blank <- read.list[(length(STD)+1):(length(STD)+2)]
    blank <- mean(blank)
    LBP.sample <- data.frame(
        sample = names,
        read = read.list[(length(STD)+3):length(read.list)]-blank,
        replicate = x
    )
})

# ---------------------Sample 41-80 -------------------------

names <- c(rep(paste0("SPL", 41:48), 2), rep(paste0("SPL", 49:56), 2), rep(paste0("SPL", 57:64), 2), rep(paste0("SPL", 65:72), 2), rep(paste0("SPL", 73:80), 2))

LBP41.80 <- list(rep1 = LBP41.80.1, rep2 = LBP41.80.2, rep3 = LBP41.80.3)

# STD tables:

LBP41.80.std <- lapply(names(LBP41.80), function(x){
    read.list <- LBP41.80[[x]] %>% 
        as.matrix() %>% 
        as.numeric()
    blank <- read.list[(length(STD)+1):(length(STD)+2)]
    blank <- mean(blank)
    LBP.std <- data.frame(
        std = STD,
        read = read.list[1:length(STD)]-blank,
        replicate = x
    )
}) %>%
    Reduce(rbind, .)

# Sample tables:

LBP41.80.sample <- lapply(names(LBP41.80), function(x){
    read.list <- LBP41.80[[x]] %>% 
        as.matrix() %>% 
        as.numeric()
    blank <- read.list[(length(STD)+1):(length(STD)+2)]
    blank <- mean(blank)

    LBP.sample <- data.frame(
        sample = names,
        read = read.list[(length(STD)+3):length(read.list)]-blank,
        replicate = x
    )
})

```

### Standard Curve

```{r std curve, echo=FALSE}
LBP1.40.std <- LBP1.40.std %>% mutate(plate = "plate1")
LBP41.80.std <- LBP41.80.std %>% mutate(plate = "plate2")

LBP.std <- rbind(LBP1.40.std, LBP41.80.std)

# standard curve:

ggplot(LBP.std, aes(read, std, color = interaction(replicate, plate))) +
    geom_point()+
    geom_smooth(method = "lm", se = F)+
    geom_vline(xintercept = c(sapply(LBP1.40.sample, function(x)max(x$read)), 
                              sapply(LBP41.80.sample, function(x)max(x$read))))
```

The correlation was not linear. The standard with a concentration of 50 µg/ml is out of the linear range, and a standard curve up to 25 µg/ml can cover the reading range of samples.

New standard curve:

```{r new std curve, echo=FALSE}
filter(LBP.std, std < 50) %>%
    ggplot(aes(read, std, color = interaction(replicate, plate))) +
    geom_point()+
    geom_smooth(method = "lm", se = F)
```

### Equation of standard curves

```{r, echo=FALSE}
LBP.std <- filter(LBP.std, std < 50)
LBP1.40.sample <- lapply(LBP1.40.sample, function(x){
    print(x$replicate[1])
    df <- filter(LBP.std, replicate==x$replicate[1] & plate == "plate1")
    mod <- lm(std~read, df)
    sum.mod <- c(mod$coefficients, "r.squared" = summary(mod)$r.squared)
    print(sum.mod)
    x$conc <- predict(mod, x)
    return(x)
})

LBP41.80.sample <- lapply(LBP41.80.sample, function(x){
    print(x$replicate[1])
    df <- filter(LBP.std, replicate==x$replicate[1] & plate == "plate2")
    mod <- lm(std~read, df)
    sum.mod <- c(mod$coefficients, "r.squared" = summary(mod)$r.squared)
    print(sum.mod)
    x$conc <- predict(mod, x)
    return(x)
})

```

### Stat

```{r, echo=FALSE}
LBP.sample <- rbind(Reduce(rbind, LBP1.40.sample), 
                    Reduce(rbind, LBP41.80.sample))


LBP.sample <- left_join(LBP.sample, metadata, by = "sample")
# LBP.sample$timepoint <- factor(LBP.sample$timepoint, levels = c("pre", "post"))

# model
mod <- lmer(conc~timepoint*treatment + (1|subject), LBP.sample)

qqnorm(residuals(mod))
qqline(residuals(mod))
shapiro.test(residuals(mod))

anova(mod, ddf = "K")

means <- emmeans(mod, specs = c("timepoint"), by = "treatment")

contrast(means, method = "pairwise")
```

The raw concentration violates the normal distribution assumption. Try square root transformation below.

```{r, echo=FALSE}
mod <- lmer(sqrt(conc)~timepoint*treatment + (1|subject), LBP.sample)

qqnorm(residuals(mod))
qqline(residuals(mod))
shapiro.test(residuals(mod))

anova(mod, ddf = "K")

means <- emmeans(mod, specs = c("timepoint"), by = "treatment")

contrast(means, method = "pairwise")
```

The QQ plot looks better after the square root transformation.

### Visualization

```{r, echo=FALSE}
df <- LBP.sample %>% 
    group_by(sample, timepoint, treatment, subject) %>%
    summarise(conc = mean(conc)) %>%
    mutate(timepoint = factor(timepoint, levels=c("pre", "post")))

ggplot(df, aes(timepoint, sqrt(conc))) +
    geom_boxplot() +
    geom_point(aes(color = subject)) +
    geom_line(aes(color = subject, group = subject)) +
    facet_wrap(~treatment) +
    theme_bw()
```
