---
title: "LBP Optimize"
author: "Eddie and Brian"
date: "2022-08-25"
output: 
        html_document:
                toc: true
                toc_float: true
---


```{r include=FALSE}
if (!require("pacman")) install.packages("pacman") ## This code installs pacman package, pacman automatically download required packages if you don't have them and load them for you.

pacman::p_load(tidyverse, here, DT, ggpubr, readxl) # These are the packages you need and to download/load


## Read Files

standard <- read_excel(here("LBP","optimize","standard","LBP_standard.xls"))

##Read files names for plasma
my_files <- list.files(path=here("LBP","optimize","tidy"),
                        pattern=".*csv")

all_csv <- lapply(here("LBP","optimize","tidy",my_files),read_csv)

# Set the name of each list element to its
# respective file name. Note full.names = FALSE to
# get only the file names, not the full path.
names(all_csv) <- gsub(".csv","",
                       list.files(here("LBP","optimize","tidy"), full.names = FALSE),
                       fixed = TRUE)

# write it into global environment.
list2env(all_csv, .GlobalEnv)

```

## Standard
```{r echo=FALSE}

standard <- standard %>%
        group_by(concentration) %>%
        summarize(absorbance = mean(absorbance_1)) 


### Linear Model
linear.model <- lm(absorbance ~ concentration, data = standard)


## Linear model function
ggplotRegression <- function (fit) {
        
        require(ggplot2)
        
        ggplot(fit$model, aes_string(x = names(fit$model)[2], y = names(fit$model)[1])) + 
                geom_point() +
                stat_smooth(method = "lm", col = "red") +
                labs(title = paste("Adj R2 = ",signif(summary(fit)$adj.r.squared, 5),
                                   "Intercept =",signif(fit$coef[[1]],5 ),
                                   " Slope =",signif(fit$coef[[2]], 5),
                                   " P =",signif(summary(fit)$coef[2,4], 5)), 
                     x = 'Concentration (ng/mL)',
                     y = 'Absorbance')
}

ggplotRegression(linear.model)

slope <- signif(linear.model$coef[[2]], 5)
intercept <- signif(linear.model$coef[[1]],5 )
```

> Need to remove higest standard

***

## Standard with high conc. removed

```{r echo=FALSE}

## Remove Highest Concntration
standard_filter <- standard %>%
        filter(!(concentration == 50.000))
    
linear.model <- lm(absorbance ~ concentration, data = standard_filter)
ggplotRegression(linear.model)

slope <- signif(linear.model$coef[[2]], 5)
intercept <- signif(linear.model$coef[[1]],5 )
```

> Better

***

## Sample CVS
```{r echo=FALSE}
plasma_dfs <- c("`1_plasma_LBP`","`1.10_plasma_LBP`","`1.100_plasma_LBP`","`1.2_plasma_LBP`")

`1_plasma_LBP` <- `1_plasma_LBP` %>%
        mutate(dilution = 0)

`1.2_plasma_LBP` <- `1.2_plasma_LBP` %>%
        mutate(dilution = 2)

`1.10_plasma_LBP` <- `1.10_plasma_LBP` %>%
        mutate(dilution = 10)

`1.100_plasma_LBP` <- `1.100_plasma_LBP` %>%
        mutate(dilution = 100)

df <- rbind(`1_plasma_LBP`,`1.2_plasma_LBP`,`1.10_plasma_LBP`,`1.100_plasma_LBP`)

## CV
df_summary <- df %>%
        group_by(sample, dilution) %>%
        summarize(mean = mean(absorbance_1),
                  sd = sd(absorbance_1),
                  cv = sd/mean *100)

datatable(df_summary, colnames = c("Sample","Dilution Factor","Absorbance","SD","CV (%)")) %>%
        formatRound(columns = c("mean","sd","cv"), digits = 3)


```

***

## Plot
```{r echo=FALSE}
##Factor
df_summary$sample <- as.factor(df_summary$sample)

ggplot(df_summary, aes(x = dilution, y = mean, color = sample)) +
        geom_point() +
        geom_line(aes(color = sample)) +
        labs(x="Dilution Factor\n (0, 2, 10, 100 Fold)", y="Absorbance",
             title = "Conclusion = Need to dilute further beyond 100 Fold")

```

*** 

## Conclusion

- Need to further dilute, you can't tell where hook effect starts.

- Eliminate highest standard

- Does not make sense to calculate LBP protein concentration with this data because it looks like the data is already within the hook effect at the 100 Fold dilution. Ideally, larger dilutions should have lower absorbance values but not the other way around. 

