---
title: "fiber_LCAT_ER"
author: "Eddie Romo"
date: '2022-08-10'
output:
  html_document:
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

```{r setup, include=FALSE}
if(!any(rownames(installed.packages()) == 'pacman'))
        install.packages('pacman')

pacman::p_load(here,
               readxl,
               tidyverse,
               lme4,
               lmerTest,
               emmeans,
               car,
               magrittr,
               dplyr,
               readxl,
               ggplot2,
               DT)



```

# ID mapping + clinic metadata

```{r echo = F}
mapping <- read.csv(here("subject_characteristics","final_samples.csv"))[,1:2] %>%
    `colnames<-`(c("Sample", "Sample.ID")) %>%
    mutate(Sample = as.character(Sample))
clinic <- read_xlsx(here("subject_characteristics","USANA Study Clinical Meta Data Sheet.xlsx"), sheet = 1) %>%
    mutate(
        Sample.ID = paste0(subject, `timepoint coded`),
        treatment = ifelse(treatment=="A", "fiber", "placebo")
    ) 
metadata <- left_join(mapping, clinic)
```

# Plate layout

```{r echo = F}
layout1 <- c(as.matrix(read_xlsx(here("LCAT","layout","LCAT_layout.xlsx"), range = "B1:L9"))) %>%
    sub(" *", "", .)
layout2 <- c(as.matrix(read_xlsx(here("LCAT","layout","LCAT_layout.xlsx"), range = "B12:L20"))) %>%
    sub(" *", "", .)

layout1 <- layout1[!is.na(layout1)]
layout1[layout1=="ercc"] <- "ercc.plate1"
layout1[layout1=="blank"] <- "blank.plate1"
layout2 <- layout2[!is.na(layout2)]
layout2[layout2=="ercc"] <- "ercc.plate2"
layout2[layout2=="blank"] <- "blank.plate2"
```

# Read and tidy LCAT assay results

```{r echo = F}
ranges <- c("C145:M153", "C157:M165", "C169:M177")

plate1 <- lapply(ranges, function(x){
    c(as.matrix(read_xlsx(here("LCAT","raw","2022-07-01_LCAT Fiber.xlsx"), range = x)))
}) %>%
    `names<-`(c("ratio1", "ratio2", "ratio3")) %>%
    do.call(cbind, .) %>%
    as.data.frame() %>%
    filter(complete.cases(.)) %>%
    mutate(Sample = layout1,
           Batch = "1") %>%
        filter(!(Sample %in% "blank.plate1")) %>%
        rowwise() %>%
        mutate(LCAT.ratio = mean(ratio1, ratio2, ratio3)) %>%
        ungroup() %>%
        mutate(norm.LCAT.ratio = LCAT.ratio/LCAT.ratio[Sample == "ercc.plate1"]) %>%
        filter(Sample != "ercc.plate1") %>%
        group_by(Sample, Batch) %>%
        summarize(LCAT.ratio = mean(LCAT.ratio),
                  norm.LCAT.ratio = mean(norm.LCAT.ratio))
        


plate2 <- lapply(ranges, function(x){
    c(as.matrix(read_xlsx(here("LCAT","raw","2022-07-02_LCAT Fiber.xlsx"), range = x)))
}) %>%
    `names<-`(c("ratio1", "ratio2", "ratio3")) %>%
    do.call(cbind, .) %>%
    as.data.frame() %>%
    filter(complete.cases(.)) %>%
    mutate(Sample = layout2, 
           Batch = "2") %>%
        filter(!(Sample %in% "blank.plate2")) %>%
        rowwise() %>%
        mutate(LCAT.ratio = mean(ratio1, ratio2, ratio3)) %>%
        ungroup() %>%
        mutate(norm.LCAT.ratio = LCAT.ratio/LCAT.ratio[Sample == "ercc.plate2"]) %>%
        filter(Sample != "ercc.plate2") %>%
        group_by(Sample, Batch) %>%
        summarize(LCAT.ratio = mean(LCAT.ratio),
                  norm.LCAT.ratio = mean(norm.LCAT.ratio))

plate.merge <- rbind(plate1, plate2) %>%
    left_join(metadata) %>%
    mutate(EU = paste(subject, timepoint, treatment, sep = ".")) %>%
        filter(Sample != "74") # Filter outlier

```

# Statistical analysis

EU=treatment:timepoint:subject.\
Linear fixed model: LCAT.ratio\~treatment\*timepoint+(1\|subject)

```{r echo = F}
## Change subject to factor
plate.merge$subject <- as.factor(plate.merge$subject)

## Reorder factor
plate.merge$timepoint <- factor(plate.merge$timepoint, levels=c('pre','post'))

# linear mixed model
lmmod <- lmerTest::lmer(norm.LCAT.ratio~treatment*timepoint + (1|subject), data = plate.merge)
# Check the assumption: normality
qqPlot(resid(lmmod))
shapiro.test(resid(lmmod))
# ANOVA
summary(lmmod)
anova(lmmod, ddf="K")
# post hoc
emm <- emmeans(lmmod, specs = c("treatment", "timepoint"), lmer.df = "K")
con <- contrast(emm, interaction = "pairwise", lmer.df = "K")
con
```

 > Results showed: (fiber.post-fiber.pre)-(placebo.post-placebo.pre)=0.0025, p-value = 0.7176\
Subjects contributed more variances than treatment and timepoint.

# Table

```{r}
datatable(plate.merge)
```

# Visulization

## All data points

```{r}
# boxplot, all data points
ggplot(plate.merge, aes(timepoint, norm.LCAT.ratio))+
    geom_boxplot()+
    geom_point(aes(color = subject), position = position_jitter())+
    # geom_line(aes(color = subject, group = subject)) +
    facet_grid(~treatment) +
    theme_bw() +
        labs(y = "Normalized \n LCAT Activity", x = "Time Point")
```

## Boxplot + lineplot

```{r}
# boxplot, mean for each subject
plate.merge %>%
    ggplot(aes(timepoint, norm.LCAT.ratio))+
    geom_boxplot()+
    geom_point(aes(color = subject))+
    geom_line(aes(color = subject, group = subject)) +
    facet_grid(~treatment) +
    theme_bw() +
        labs(y = "Normalized \n LCAT Activity", x = "Time Point")
```

## Mean for each subject + batch 1&2

```{r}
# boxplot, show batch 1&2
plate.merge %>%
    group_by(subject, timepoint, treatment, Batch) %>%
    ggplot(aes(timepoint, norm.LCAT.ratio))+
    geom_boxplot()+
    geom_point(aes(color = subject))+
    geom_line(aes(color = subject, group = subject)) +
    facet_grid(Batch~treatment) +
    theme_bw() +
        labs(y = "Normalized \n LCAT Activity", x = "Time Point")
```

## HDL-C correlation with LCAT
```{r}
# CHeck normality
shapiro.test(plate.merge$norm.LCAT.ratio)
shapiro.test(plate.merge$HDL)

## Log transform HDL

plate.merge <- plate.merge %>%
        mutate(log_HDL = log(HDL))

shapiro.test(plate.merge$log_HDL)

cor.test(plate.merge$norm.LCAT.ratio, plate.merge$log_HDL, method = "pearson")
```

