---
title: "LCAT microbiome correlation"
author: "Cynthia Tang, modfied 9.28.22 BH"
date: "2022-09-24"
output:
  html_document:
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

if(!any(rownames(installed.packages()) == 'pacman'))
        install.packages('pacman')

if(!any(rownames(installed.packages()) == 'devtools'))
        install.packages('devtools')

if(!any(rownames(installed.packages()) == 'HTSet'))
        devtools::install_github("zhuchcn/HTSet")

pacman::p_load(HTSet,
               tidyverse,
               readxl,
               DT,
               insight,
               here)
```

# QC summary

> Earlier QC analysis suggest plate effects in which the overall fluorescence intensity in plate 2 was significantly lower than plate 1. Therefore, the subsequent analysis uses **Normalized LCAT activity**.

> LCAT activity was normalized to quality control (ERCC2 plasma) by taking the ratio of the Sample LCAT activity (390nm/470nm) / ERCC2 Plasma LCAT activity (390nm/470nm). 

> Sample #73 or subject 223B was a clear outlier, likely due to technical error, and removed from correlation analysis. If there is another partial kit avaiable, I would suggest rerunning sample #73 along with an ERCC2 plasma as QC.

***

## Load Data
```{r}
plate.merge <- readRDS(here("LCAT","data","LCAT_summary_filter_223B.rds"))
```

# OTU table

At Genus level:
```{r}
taxa <- readRDS(here("microbiome","data","taxa.rds"))

overlapped.id <- sampleNames(taxa) %in% plate.merge$Sample.ID

genus <- summarize_feature(taxa, "genus")[,overlapped.id]

genus <- genus$edata # OTU count table at the genus level

var0 <- apply(genus, 1, function(x)var(x)==0) # remove genus with variance == 0

sum(var0)

## Remove Outlier (Sample # 74 / Sample ID 223B)

genus <- genus[, colnames(genus)!="223B"]


```

# Correlation analysis

Pull Normalized LCAT ratios as vector.
```{r}
LCAT <- plate.merge %>%
    select(Sample.ID, norm.LCAT.ratio) %>%
    pull() %>%
    `names<-`(unique(plate.merge$Sample.ID))

# Check whether the sample ID in the two dataset matches
identical(names(LCAT), colnames(genus))
```

Correlation between absolute microbes (genus) OTU counts and LCAT concentration.
```{r echo = F, warning = F}
correlation <- function(microbes, y, cor.method){
    # correlation
    cor <- apply(microbes, 1, function(x){
        cor.test(x, y, method = cor.method, exact = F)$estimate
    })
    # p-value
    p <- apply(microbes, 1, function(x){
        cor.test(x, y, method = cor.method, exact = F)$p.value
    })
    # adjusted p-value
    padj <- p.adjust(p, method = "BH")
    data.frame(
        cor = cor,
        pval = p,
        padj = padj
    )
}


cor.count <- correlation(microbes = genus, y = LCAT, cor.method = "spearman") %>%
        mutate(p.value.star = insight::format_p(padj, name = NULL, stars_only = TRUE))
```

***

# Visualization Function
```{r function, echo = F}
top.hits <- filter(cor.count, padj<0.05) %>%
    arrange(desc(abs(cor))) %>%
    rownames()

cor.plot <- function(microbes, x, y, xlab = x, ylab = "LCAT", cor.res = cor.count){
    df <- data.frame(
        x = microbes[x,],
        y = y
    ) 
    stats <- as.numeric(cor.count[x,c(1,3)]) %>%
        signif(3)
    p <- ggplot(df, aes(x=x, y=y)) +
        geom_point() +
        geom_smooth(method = "lm") +
        labs(x = paste(xlab, "count"), y = ylab) +
        theme_bw()
    xmax <- max(df$x)
    ymax <- max(df$y)
    p <- p +
        annotate(
            "text", x = xmax*0.8, y = c(ymax, ymax*0.98), 
            label = paste0(c("rho = ", "padj = "), stats)
        )
    return(p)
}
```

***

## Correlation plot with Bifidobacterium Adjusted Pvalue 
```{r echo = F}
cor.plot(microbes = genus, x = "Bifidobacterium", y = LCAT, ylab = "LCAT", cor.res = cor.count) +
    scale_x_log10()
```

***

## All Correlation plot for Adjusted Pvalue < 0.05
```{r echo = F, warning = F}

for(i in top.hits){
    plots <- cor.plot(microbes = genus, x = i, y = LCAT, ylab = "LCAT", cor.res = cor.count)  +
    scale_x_log10()
    print(plots)
}

```


***

## Table of Correlation
```{r}
datatable(cor.count, colnames = c("Genus","Spearman Correlation (rho)","p-value","Adjusted p-value","Sig level")) %>%
        formatRound(columns = "cor", digits = 2) %>%
        formatRound(columns = c("pval","padj"), digits = 3)
```

