---
title: "LBP Optim Trial 2"
author: "Eddie Romo"
date: "2022-08-25"
output: 
        html_document:
                toc: true
                toc_float: true
---


```{r include=FALSE}
if (!require("pacman")) install.packages("pacman") ## This code installs pacman package, pacman automatically download required packages if you don't have them and load them for you.

pacman::p_load(tidyverse, here, DT, ggpubr, readxl) # These are the packages you need and to download/load


## Read Files

standard <- read_excel(here("LBP","optimize_2","standard","LBP_standard.xls"))

##Read files names for plasma
my_files <- list.files(path=here("LBP","optimize_2","tidy"),
                        pattern=".*csv")

all_csv <- lapply(here("LBP","optimize_2","tidy",my_files),read_csv)

# Set the name of each list element to its
# respective file name. Note full.names = FALSE to
# get only the file names, not the full path.
names(all_csv) <- gsub(".csv","",
                       list.files(here("LBP","optimize_2","tidy"), full.names = FALSE),
                       fixed = TRUE)

# write it into global environment.
list2env(all_csv, .GlobalEnv)

```

## Standard
```{r echo=FALSE}

standard <- standard %>%
        group_by(concentration) %>%
        summarize(absorbance = mean(absorbance_1)) 


### Linear Model
linear.model <- lm(absorbance ~ concentration, data = standard)


## Linear model function
ggplotRegression <- function (fit) {
        
        require(ggplot2)
        
        ggplot(fit$model, aes_string(x = names(fit$model)[2], y = names(fit$model)[1])) + 
                geom_point() +
                stat_smooth(method = "lm", col = "red") +
                labs(title = paste("Adj R2 = ",signif(summary(fit)$adj.r.squared, 5),
                                   "Intercept =",signif(fit$coef[[1]],5 ),
                                   " Slope =",signif(fit$coef[[2]], 5),
                                   " P =",signif(summary(fit)$coef[2,4], 5)), 
                     x = 'Concentration (ng/mL)',
                     y = 'Absorbance')
}

ggplotRegression(linear.model)

slope <- signif(linear.model$coef[[2]], 5)
intercept <- signif(linear.model$coef[[1]],5 )
```

> removal of highest concentration.

***

## Standard with high conc. removed

```{r echo=FALSE}

## Remove Highest Concntration
standard_filter <- standard %>%
        filter(!(concentration == 50.000))
    
linear.model <- lm(absorbance ~ concentration, data = standard_filter)
ggplotRegression(linear.model)

slope <- signif(linear.model$coef[[2]], 5)
intercept <- signif(linear.model$coef[[1]],5 )
```

> significantly improved R2

***

## Sample CVS
```{r echo=FALSE}
plasma_dfs <- c("`1.100_plasma_LBP`","`1.200_plasma_LBP`","`1.400_plasma_LBP`","`1.800_plasma_LBP`","`1.1600_plasma_LBP`")

`1.100_plasma_LBP` <- `1.100_plasma_LBP` %>%
        mutate(dilution = 100)

`1.200_plasma_LBP` <- `1.200_plasma_LBP` %>%
        mutate(dilution = 200)

`1.400_plasma_LBP` <- `1.400_plasma_LBP` %>%
        mutate(dilution = 400)

`1.800_plasma_LBP` <- `1.800_plasma_LBP` %>%
        mutate(dilution = 800)

`1.1600_plasma_LBP` <- `1.1600_plasma_LBP` %>%
        mutate(dilution = 1600)

df <- rbind(`1.100_plasma_LBP`,`1.200_plasma_LBP`,`1.400_plasma_LBP`,`1.800_plasma_LBP`,`1.1600_plasma_LBP` )

## CV
df_summary <- df %>%
        group_by(sample, dilution) %>%
        summarize(mean = mean(absorbance_1),
                  sd = sd(absorbance_1),
                  cv = sd/mean *100)

datatable(df_summary, colnames = c("Sample","Dilution Factor","Absorbance","SD","CV (%)")) %>%
        formatRound(columns = c("mean","sd","cv"), digits = 3)


```

***

## Plot
```{r echo=FALSE}
##Factor
df_summary$sample <- as.factor(df_summary$sample)

ggplot(df_summary, aes(x = dilution, y = mean, color = sample)) +
        geom_point() +
        geom_line(aes(color = sample)) +
        labs(x="Dilution Factor\n (100, 200, 400, 800, 1600 Fold)", y="Absorbance",
             title = "Conclusion: oversaturation solved.")

```

*** 

## Conclusion

- The most dilute sample 1:100 appeared to be within the range of saturation from earlier trial run where signal would decrease with increase concentration. Here after 1:100 the absorbance begins to decrease with higher concentration as expected.

- Additional trial run appears to have solved hook effect phenomenon with higher dilution factors.

- 1.800 dilution optimal dilution factor for Fiber Study samples. 

- Eliminate highest standard


